---
title: "Exploratory Data Analysis Course Project 2"
author: "Eric Tiongson"
date: "2/21/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

## Introductions

### Overview

This paper was created as a project of the Exploratory Data Analysis course in Coursera.

As stated in the project description:

> The overall goal of this assignment is to explore the National Emissions Inventory database and see what it say about fine particulate matter pollution in the United states over the 10-year period 1999–2008. 

### Course Data

Fine particulate matter (PM2.5) is an ambient air pollutant for which there is strong evidence that it is harmful to human health. In the United States, the Environmental Protection Agency (EPA) is tasked with setting national ambient air quality standards for fine PM and for tracking the emissions of this pollutant into the atmosphere. Approximatly every 3 years, the EPA releases its database on emissions of PM2.5. This database is known as the National Emissions Inventory (NEI). 

Additional information about the NEI can be found at the [EPA National Emissions Inventory web site](http://www.epa.gov/ttn/chief/eiinformation.html).

The NEI project data was downloaded and loaded into the environment as shown below.

```{r load-data, cache=TRUE}
exdata_filename <- "exdata-data-NEI_data.zip"

if (!file.exists(exdata_filename)) {
  download_url <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip"
  download.file(download_url, destfile = exdata_filename)
  unzip (zipfile = exdata_filename)
}

if (!exists("NEI")) {
  # print("Loading NEI Data, please wait.")
  NEI <- readRDS("summarySCC_PM25.rds") 
}

if (!exists("SCC")) {
  # print("Loading SCC Data.")
  SCC <- readRDS("Source_Classification_Code.rds")
}
```

*PM2.5 Emissions Data* (summarySCC_PM25.rds) contains a data frame with all of the PM2.5 emissions data for 1999, 2002, 2005, and 2008. For each year, the table contains number of tons of PM2.5 emitted from a specific type of source for the entire year. Here are the first few rows.

```{r show-summary}
head(NEI)
```

The PM2.5 variables are as follows:

* fips: A five-digit number (represented as a string) indicating the U.S. county
* SCC: The name of the source as indicated by a digit string (see source code classification table)
* Pollutant: A string indicating the pollutant
* Emissions: Amount of PM2.5 emitted, in tons
* type: The type of source (point, non-point, on-road, or non-road)
* year: The year of emissions recorded

*The Source Classification Table* (Source_Classification_Code.rds) provides a mapping from the SCC digit strings in the Emissions table to the actual name of the PM2.5 source.

```{r show-classification}
head(SCC[,c("SCC", "Short.Name")])
```

## Questions and Answers

The project aims to answer the questions listed below.  The answers to the questions are illustrated in the plots following each question.

**Question #1**: Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.
```{r total-annual-emissions, cache=TRUE}
total_annual_emissions <- aggregate(Emissions ~ year, NEI, FUN = sum)
color_range <- 2:(length(total_annual_emissions$year)+1)
with(total_annual_emissions, 
     barplot(height=Emissions/1000, names.arg = year, col = color_range, 
             xlab = "Year", ylab = expression('PM'[2.5]*' in Kilotons'),
             main = expression('Annual Emission PM'[2.5]*' from 1999 to 2008')))
```

**Question #2**: Have total emissions from PM2.5 decreased in the Baltimore City, Maryland ( fips == "24510") from 1999 to 2008? Use the base plotting system to make a plot answering this question.

```{r baltimore-emissions}
b_total_emissions <- NEI %>%
  filter(fips == "24510") %>%
  group_by(year) %>%
  summarise(Emissions = sum(Emissions))

with(b_total_emissions,
       barplot(height=Emissions/1000, names.arg = year, col = color_range, 
               xlab = "Year", ylab = expression('PM'[2.5]*' in Kilotons'),
               main = expression('Baltimore, Maryland Emissions from 1999 to 2008'))
)
```

**Question #3**: Of the four types of sources indicated by the  (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 19992008 for Baltimore City? Which have seen increases in emissions from 19992008? Use the ggplot2 plotting system to make a plot answer this question.

```{r baltimore-emissions-by-source}
b_emissions <- NEI %>%
  filter(fips == "24510") %>%
  group_by(year, type) %>%
  summarise(Emissions=sum(Emissions))

b_em_plot <- ggplot(data = b_emissions, aes(x = year, y = Emissions, color = type)) +
  geom_line(size = 2) + xlab("Year") + ylab(expression('PM'[2.5]*' Emission')) +
  ggtitle("Baltimore Emissions by Source Type")
print(b_em_plot)
```

*Question #4**: Across the United States, how have emissions from coal combustion-related sources changed from 1999 to 2008?

```{r coal-emissions}
coal_SCC <- SCC[grep("[Cc][Oo][Aa][Ll]", SCC$EI.Sector), "SCC"]
coal_NEI <- NEI %>% filter(SCC %in% coal_SCC)
coal_summary <- coal_NEI %>% group_by(year) %>% summarise(Emissions = sum(Emissions))

c_plot <- ggplot(coal_summary, aes(x=year, y=round(Emissions/1000,2), label=round(Emissions/1000,2), fill=year)) +
  geom_bar(stat="identity") + ylab(expression('PM'[2.5]*' Emissions in Kilotons')) + xlab("Year") +
  geom_label(aes(fill = year),colour = "white", fontface = "bold") +
  ggtitle("Coal Combustion Emissions, 1999 to 2008.")
print(c_plot)
```

*Question #5**: How have emissions from motor vehicle sources changed from 19992008 in Baltimore City?

* Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California ( == ""). Which city has seen greater changes over time in motor vehicle emissions?
